Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2018-01-15 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Text updates

\begin{document}
\maketitle
\begin{abstract}
\singlespacing
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to 1) apply standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
\end{abstract}

@@

\section{Introduction} 
%% Sequencing technologies
Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs. As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}. 
%% Rare variant methods
In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies of each rare variant. Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants. When designing an `aggregation' method, there are three questions that are usually considered. First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}? Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly. In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.

%% The model comparison method
In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests. The MRP framework, as implemented, can exploit correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).

In Bayesian model comparison, we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.

%% Use of summary statistic data
In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.  We present a derivation where summary statistics, such as estimates of effect size and the corresponding standard error from single variant and single phenotype analysis, are the input data. Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.

%% Genome variant annotation and protein truncating variants
Genome variant annotation is critical for the interpretation of genetic findings. In the MRP model comparison we present approaches for including priors on spread of effect sizes. An example of an important class of variants where functional prediction is much more straightforward is the class of variants predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.

%% Protective modifiers of disease risk
Given the translational potential of identifying protective modifiers of human disease risk we characterize how model comparison may be used to improve discovery of such protective signals. We present an extension allowing for modeling of location (direction) of genetic effects, which may be useful for prioritizing variants or genes that are consistent with a protective profile of disease\cite{pcsk9,cohen2006sequence,sullivan2012effect}.

@@

%%Apply methods to the GLGC data and exome sequencing data sets. 
We also apply MRP to two groups of related phenotypes. First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset. We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265). We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.

@@

\noindent When we have data from multiple studies 
we allow for possible differences in genetic effects across ethnicities or populations 
extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
and the summary statistics approach of RAREMETAL ~\cite{liu2014meta} from univariate to multivariate phenotypes.
Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes. 
As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
Thus, the prior is
$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$

>>>>

\begin{document}
\maketitle
\begin{abstract}
\singlespacing
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
\end{abstract}

@@

\section{Introduction} 
%% Sequencing technologies
Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs. As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}.

%% Rare variant methods
In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies of each rare variant. Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants. When designing an `aggregation' method, there are three questions that are usually considered. First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}? Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to use gene discovery approaches that leverage the information from multiple phenotypes jointly. In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.

%% The model comparison method
In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests. The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}). MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered. The BF is an alternative to $p$-values from traditional hypothesis testing. For MRP, the BF represents the statistical evidence for a non-zero effect for a particular group of rare variants on the phenotype(s) of interest.

%% Use of summary statistic data
While many large genetic consortia generally collect both raw genotype and phenotype data, in practice, sharing of individual genotype and phenotype data across groups is difficult to achieve. To address this, MRP can take summary statistics, such as estimates of effect size and the corresponding standard error from typical single variant-single phenotype linear or logistic regressions, as input data. Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.

%% Genome variant annotation and protein truncating variants and  protective modifiers of disease risk
Aggregation techniques rely on variant annotations to assign variants to groups for analysis. MRP allows for the inclusion of priors on the spread of effect sizes that can be adjusted depending on what type of variants are included in the analysis. For instance, protein truncating variants (PTVs)\cite{rivas2013assessing,rivas2015effect} are an important class of variants that are more likely to be functional because they often disrupt the normal function of a gene. This biological knowledge can be reflected in the choices of priors for PTVs in MRP. Since PTVs typically abolish gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}. We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.

@@

%%Apply methods to the GLGC data and exome sequencing data sets. 
We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank. First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062). We find that XXX. We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find strong evidence that rare PTVs in \textit{ANGPTL7} protect against glaucoma. These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.

@@

\noindent When we have data from multiple studies 
we allow for possible differences in genetic effects across ethnicities or populations 
extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
and the summary statistics approach of RAREMETAL~\cite{liu2014meta} from univariate to multivariate phenotypes.
Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes. 
As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
Thus, the prior is
$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$
